# IDENTIFICATION AND CHARACTERIZATION OF HEAVY METAL TOLERANCE OF BACTERIA ISOLATED FROM COAL ASH {#metal}

## Introduction

Coal ash is a waste product generated from coal-fired power plants. These power plants are located near water sources to meet water demands during the electricity generation process. The ash and other waste products are discharged as a slurry into a settling pond, after the combustion process [@connorsCoalashManagementElectric2015]. To prevent overflow the pond is maintained by pumping surface water into the nearby waters, usually a river. In North Carolina, 14 such coal ash ponds are maintained by Duke Energy.  Unlined and uncapped coal ash waste ponds therefore have the potential for leaching contaminants into the groundwater or spilling into nearby waterways possibly polluting drinking water and/or disrupting aquatic ecological communities [@otterTrophicStatusMetal2012; @ruhlSurveyPotentialEnvironmental2009].    

On February 2, 2014, a coal ash pond located at the retired Dan River Steam Station near Eden, NC expelled approximately 39,000 tons of coal ash slurry into the Dan River of North Carolina due to an underground pipe collapse [@ncdeqRapidResponseTimeline2014]. The ash was dispersed along the river bottom between the spill site and the Kerr Reservoir, approximately 120 km downstream [@hesterbergAssessmentTraceelementImpacts2014]. Dredging operations conducted by Duke Energy at Town Creek and an area upstream from the Schoolfield Dam in Danville, VA, removed approximately 2,500 tons of coal ash-laden sediment, but the remainder was deposited in the river bottom [@lemlyDamageCostDan2015].

Physical and chemical properties of coal ash are determined by the geographical location where the raw coal was mined, the type of boiler, and the operating conditions of the power plant [@jayaranjanReuseOptionsCoal2014].  All natural elements can be found in coal ash including trace elements such as arsenic, cadmium, chromium, mercury, lead, selenium, and zinc [@greelyjr.EffectsSedimentContaining2014; @jayaranjanReuseOptionsCoal2014].  Coal ash is wholly comprised of multiple waste products with differing chemical properties including fly ash, bottom ash, and byproducts of pollution mitigation processes, such as scrubber systems. Fly ash is composed mainly of oxides such as, $\mathrm{SiO_2}$, $\mathrm{Al_2O}$, $\mathrm{Fe_2O}$. Bottom ash consists of silicate, carbonate, aluminate, ferrous materials and high concentrations of several heavy metals and metalloids [@kiskuCharacterizationCoalFly2018]. 

Heavy metals are characterized by a density greater than 5 g/$\mathrm{cm^3}$, mostly transition elements, and play an important role as trace elements in biochemical reactions. These heavy metals, due to an incompletely filled d orbital, allow the cations to form complex compounds with the potential to be redox reactive. Heavy metal ions form unspecific complex compounds in living cells, leading to toxic effects. For example, $\mathrm{Hg^{2+}}$, $\mathrm{Cd^{2+}}$, and $\mathrm{Ag^{+}}$ form strong toxic complexes that are not conducive to any physiological function and are highly toxic. Trace metals such as $\mathrm{Zn^{2+}}$, $\mathrm{Ni^{2+}}$, and $\mathrm{Cu^{2+}}$ are required for some biological functions but they are are toxic at high concentrations @hesterbergAssessmentTraceelementImpacts2014. Therefore, the intracellular concentration of heavy metals must be controlled, and organisms have adapted heavy metal resistance strategies [@binkleyHeavyMetalsWastewater2003; @poulainCrackingMercuryMethylation2013].  

Through the process of coal combustion, the coal ash is rendered sterile. Therefore, inoculation of coal ash in waste ponds must occur by natural processes including atmospheric deposition through rainfall and windblown particulates. Opportunistic microorganisms may be adapted, or may be able to adapt, to the high concentration of heavy metals through resistance or metal detoxification. In addition to naturally occurring processes, anthropogenic perturbations such as coal ash spills result in altered environments in receiving waters, which microbes may adapt to and may mobilize coal ash constituents into food webs. Knowledge of the microbial community structure and distribution is important to estimate the extent of biological mobility of these pollutants. Further, organisms which exhibit metal tolerance may warrant further investigation into their potential for bioremediation [@keshriProfileBacterialCommunities2014; @naikLeadResistantBacteria2013; @pepiMercuryresistantBacterialStrains2011; @rajaCharacterizationBoronTolerant2012].

The objective of this study was to determine if a microbial community is viable and present in coal ash ponds, and if present, identify taxa, and assess the metal tolerance of isolated organisms. Previous studies have analyzed the microbial community of soils  amended with coal ash[@klubekCharacterizationMicrobialAbundance1992]. No studies to date have determined the microbial community of coal ash alone. Through preliminary PCR analysis, a small amount of DNA was present in a coal ash sample from the Dan River Steam Station coal ash pond, but we were not able to amplify SSU rDNA from bacteria with universal prokaryotic primers, possibly due to contaminants present in the sample or low concentrations of bacterial DNA. Therefore, in this study, we increased the abundance of microbes and their DNA by culturing and isolating bacteria from samples of coal ash, to ensure adequate amounts of DNA for analyses. 

### Objectives and hypotheses

**Objective 1: To identify isolated bacteria using qPCR amplification and rDNA sequencing. I seek to identify these taxa using qPCR amplification and rDNA sequencing to compare to the GenBank database.**

*I hypothesize novel organisms may be discovered which may have unique metal tolerance capabilities that may be useful for bioremediation.*

**Objective 2: To determine the metal tolerance of bacteria isolated from coal ash collected from the Dan River impoundment site.**

*I hypothesize that microbial growth of bacterial isolates from pure coal ash collected at the Dan River Steam Station impoundment pond will be more tolerant to heavy metals than compared to organisms isolated from a reference site (soil collected from Peabody Park on UNCG campus).*

## Methods
### Pure culture isolation from coal ash
Samples of coal ash provided by our collaborator, Brian Williams of the Dan River Basin Association were taken directly from a coal ash retention pond at the retired Dan River Steam Station near Eden, NC. To culture isolated organisms, aliquots (0.5 g) of coal ash from the pond were added to six 50 mL conical tubes. 40 mL of filter sterilized (0.22 µm pore) Dan River water was added to three tubes and filter sterilized Dan River water supplemented with 10% nutrient broth to the other three tubes.  Additionally, filter sterilized Dan River water alone was evaluated in triplicate to serve as a sterility control. Tubes were incubated at ambient room temperature for 48 hours without agitation. 1 mL aliquots were transferred from each culture to 50% nutrient agar plates for isolation . Morphologically unique colonies identified on spread plates were isolated and pure cultures maintained on 50% nutrient agar slants for heavy metal tolerance experimentation. Sediment samples collected from Peabody Park, UNCG campus, used as a comparison, were isolated using the same protocol as described. 

### Heavy metal characterization
Optical density (OD) at 580 nm, of liquid broth culture was assessed as a proxy for growth of organisms, since growth is related to the increase in turbidity of a bacterial culture.  Metals tested include, arsenic, cadmium, chromium, mercury, lead, selenium and zinc. Stock concentrations of $\mathrm{Na_2HAsO_4}$, $\mathrm{CdCl_2}$, $\mathrm{CrCl_3}$, $\mathrm{HgCl_2}$, $\mathrm{PbCl_2}$, $\mathrm{Na_2SeO_3}$, and $\mathrm{ZnCl_2}$ were serially diluted to span at least two orders of magnitude greater and less than natural environmental concentrations reported for each metal, resulting in 5 concentration levels per metal (Table \@ref(tab:metals)). 

```{r metals, echo=FALSE, fig.cap="caption"}
library(readxl)
library(kableExtra)
library(tidyverse)
options(scipen=999)
read_excel("data/pcrdata.xlsx", sheet = "metal_conc") %>% 
  kable(booktabs=T, caption = "Final concentrations (µg/L) of metal in fermentation broth.", align = "c" ) 
```

Metal stocks were prepared by dissolving each metal salt in sterile reverse osmosis deionized (RO/DI) water and subsequently diluted. To ensure a viable culture inoculum, sufficient volume of cells, and standardization for experimentation, each unique isolate was incubated in a 50% nutrient broth culture in a shaking water bath at 25ºC overnight until an optical density of 0.7 at 580 nm was observed. For each metal-isolate experiment, 100 µL of 0.7 OD~580~ nm inoculum was added to 2.4 mL of 50% nutrient broth amended with 100 µL each concentration of metal in triplicate. For the control, 100 µL of RO/DI water was substituted for the metal spike in triplicate. Growth patterns were measured by absorbance at 580 nm, and recorded at nine time points, approximately 0, 1, 2, 4, 8, 16, 24, 30, 36, and 48 hours post inoculation. 

### Isolate identification and phylogenetic analysis 
Isolates were grown in 50% nutrient broth then pelleted with centrifugation. The pellet was resuspended CTAB buffer. DNA was extracted and purified using the CTAB method [@stewartRapidCTABDNA1993], then amplified with universal 16S primers [@bruceAmplificationDNANative1992; @edwardsIsolationDirectComplete1989]. PCR products were then purified commercially sequenced. Sequence chromatogram data was evaluated and edited manually to optimize the sequence. The FASTA files of sequences were imported into Molecular Evolutionary Genetics Analysis version 7.0 (MEGA7) [@kumarMEGA7MolecularEvolutionary2016]. Sequences were aligned using MUSCLE and manual adjustments [@edgarMUSCLEMultipleSequence2004]. The resulting aligned sequences were then subjected to phylogenetic tree analysis using MEGA7. The maximum likelihood tree was computed using MEGA7 using the Jukes-Cantor model [@jukesEvolutionProteinMolecules1969]. The bootstrap consensus tree was inferred from 1000 replicates. Initial trees with a greater log likelihood value were calculated by applying Neighbor-Join and BioNJ algorithms to a matrix of pairwise distances estimated using the Maximum Composite Likelihood approach. 

To identify the genus of each isolate, the sequences were compared to the NCBI GenBank database with BLASTn (Nucleotide Basic Local Alignment Search Tool). Genus calls were made with similarities greater than 97% [@clarkGenBank2016].

### Statistical analyses

To characterize the heavy metal tolerance of isolates in liquid broth culture, the carrying capacity and growth rate of each isolate-metal experiment was modeled with a Gompertz sigmoidal function [@zwieteringModelingBacterialGrowth1990]. Significant differences of growth rate and carrying capacity within each experiment were determined using the Kruskall-Wallis one-way analysis of variance test (p-value < 0.05). Experiments were excluded from analysis where no growth was observed in the control. Isolates were deemed metal tolerant at the concentrations tested if no significant difference was observed between concentrations. 

## Results
### Phylogenetic analysis
A total of 31 unknown microorganisms were isolated from the coal ash sample and sequenced, five were isolated from Peabody Park soil for comparison to coal ash samples to "control" samples. The soil samples are assumed to be free of coal ash, but these soils were probably affected by farming practices in the past, which may include introduction of heavy metals to the soil. Sequences were aligned and a phylogenetic tree was constructed (Figure \@ref(fig:tree)). Sequences were predominantly identified as *Bacillus* and *Arthrobacter* spp. One isolate, DR 76, was not identified in GenBank with the BLASTn search. Of the 5 isolates that were cultured from soil, three are identified as *Bacillus* sp., one *Duganella*, and one *Pseudomonas*. *Duganella* and *Pseudomonas* were unique to the soil samples. 

### Metal Tolerance assays
We tested metal tolerance of 16 isolates. When exposed to a heavy metal, isolates generally grew well (Figures \@ref(fig:si4) - \@ref(fig:dr52)). The isolates were challenged against seven metals, resulting in a total of 94 trials. 17 trials were excluded due to lack of growth. There was little difference between coal ash isolates and a "control", SI 4 (Figure \@ref(fig:si4)). In one of the "control" isolates, SI 2, growth was not observed and was excluded. The soil isolate SI 4, was tolerant to all metals at the concentrations tested except for cadmium and zinc (Tables \@ref(tab:kpvalue) and \@ref(tab:muvalue), Figure \@ref(fig:si4). 

Of the growth curves analyzed, when evaluating carrying capacity, 70 trials are not different from the control (Kruskall-Wallis test p-value >0.05). (Table \@ref(tab:kpvalue)). Evaluation of maximal growth rate produces 71 isolate-metal trials that are not significantly different (Table \@ref(tab:muvalue)). 64% of the isolates tested are susceptible to cadmium and 72% are susceptible to selenium. The other metals tested (As, Cr, Hg, Pb, and Zn) resulted in much less susceptibility, ranging from 1-3 significant hits per metal. 


\begin{center}
  \includegraphics[width=400pt]{figure/tree}
  \captionof{figure}[Phylogenetic analysis of coal ash isolates.]{Phylogenetic analysis of isolates cultured from coal ash amended media. Identification of Coal ash-free isolates are prepended with S1-S5, all others are coal ash samples.}\label{fig:tree}
\end{center}

```{r}
kp <- read.csv("data/kpvalues.csv",stringsAsFactors=FALSE )  %>%  mutate_if(is.numeric, round, digits=3) #%>% replace(is.na(.), 0) 

mup <- read.csv("data/muvalues.csv",stringsAsFactors=FALSE )  %>%  mutate_if(is.numeric, round, digits=3) #%>% replace(is.na(.), 0) 
```


``` {r kpvalue}
library(kableExtra)
library(dplyr)

options(kableExtra.latex.bsTable = T)
kp %>%
  mutate_if(is.numeric, function(x) {
    cell_spec(x, bold = ifelse(x < 0.05, T, F))
  }) %>%
  kable(escape = F, align = "l", linesep = "", booktabs = T, caption.short = "P-values of Kruskall-Wallis test of carrying capacity.",
        caption = "P-values of Kruskall-Wallis test of carrying capacity. Bold typeface signifies p-value <0.05.") %>%
  row_spec(0, align = "c") %>% 
  kable_styling(latex_options = "striped",  
                full_width = F)%>%
  kable_styling(latex_options = "scale_down")

```

``` {r muvalue}
options(kableExtra.latex.bsTable = T)
mup %>%
  mutate_if(is.numeric, function(x) {
    cell_spec(x, bold = ifelse(x < 0.05, T, F))
  }) %>%
  kable(escape = F, align = "l", linesep = "", booktabs = T, caption.short = "P-values of Kruskall-Wallis test of maximal growth rate.",
        caption = "P-values of Kruskall-Wallis test of maximal growth rate. Bold typeface signifies p-value <0.05.") %>%
  row_spec(0, align = "c") %>% 
  kable_styling(latex_options = "striped", 
                full_width = F)%>%
  kable_styling(latex_options = "scale_down")

```
\begin{figure}
  \caption[Growth of soil isolate SI4.]{Growth of soil isolate SI4. Lines connect means of three replicates. Carrying capacity is not significantly different between metal concentrations. Growth rates are different in Cd and Zn experiments (Kruskall-Wallis test p-value < 0.05)}\label{fig:si4}
  \centering
    \includegraphics[width=400pt]{figure/si4plot}
\end{figure}

\begin{figure}
  \caption[Growth of isolate DR13.]{Growth of isolate DR13. Lines connect means of three replicates. Carrying capacity is significantly different between metal concentrations of Cd and Zn.  Growth rates are different with As and Zn (Kruskall-Wallis test p-value < 0.05)}\label{fig:dr13}
  \centering
    \includegraphics[width=400pt]{figure/dr13plot}
\end{figure}

\begin{figure}
  \caption[Growth of isolate DR3.2.]{Growth of isolate DR3.2. Lines connect means of three replicates. Carrying capacity is significantly different between metal concentrations of Cd and Se. No difference is observed when testing maximal growth rates (Kruskall-Wallis test p-value < 0.05)}\label{fig:dr32}
  \centering
    \includegraphics[width=400pt]{figure/dr32plot}
\end{figure}

\begin{figure}
  \caption[Growth of isolate DR52.]{Growth of isolate DR52. Lines connect means of three replicates. Se was excluded from analysis. Carrying capacity is significantly different between metal concentrations of Cd. Growth rates are different with As, Cd, and Hg. (Kruskall-Wallis test p-value < 0.05) }\label{fig:dr52}
  \centering
    \includegraphics[width=400pt]{figure/dr52plot}
\end{figure}
  
## Discussion
We isolated cultures of aerobic bacteria from samples of coal ash. The coal ash isolates are predominantly identified as *Arthrobacter* and *Bacillus* spp. *Arthrobacter* spp. are ubiquitous and have been found in common soils as well as extreme environments. These spore-formers are able to survive long periods of time under stressful conditions such as nutrient deficiency, temperature shifts, and toxic chemicals. *Arthrobacter* sp. are metabolically diverse and have been utilized to biodegrade environmental pollutants and are highly resistant to heavy metals [@mongodinSecretsSoilSurvival2006]. We isolated  *Arthrobacter* from coal ash samples, but did not isolate from the soil sample. The predominance of *Arthrobacter* sp. from the coal ash may be due to the selective pressure of the coal ash, although our culturing method could have also selected for these genera. High concentrations of cadmium has an effect on most of the isolates tested. Of the 5 isolates that are tolerant of Cd, one is *Arthrobacter*, although not all *Arthrobacter* are tolerant of Cd. *Bacillus* spp. are genomically diverse, ubiquitous endospore-formers are highly resistant to stressors [@earlEcologyGenomicsBacillus2008]. Three *Bacillus* spp. were subjected to metal tolerance assays. The soil isolate SI2, a *Bacillus* was only tested for Se, and was susceptible. The other *Bacillus* spp. (DR52 and DRp51) exhibited the same tolerances to the metals; tolerant to all except Cd and Se. 

One isolate could not be identified by a BLASTn search of the GenBank database, and may be an interesting organism to for further testing and characterization. Since we only studied organisms which were easily grown and maintained within the lab under room temperature conditions, only a small portion of the organisms present are cultivated. More novel organisms may be discoverable using less selective methods.   

These data are suggestive of metal tolerance, this assay is useful as a screen to find interesting isolates for further study, but these data are not wholly conclusive. Due to a small sample size and unbalanced number of coal ash isolates to soil isolates, we are unable to determine if the soil samples are more or less tolerant to heavy metals than the coal ash samples. Although, with these methods, when comparing isolates from soil devoid of coal ash contamination to coal ash isolates, there seem to be no obvious differences. 

